Real-space computation of $E$/$B$-mode maps I: Formalism, Compact Kernels, and Polarized Filaments

### 2018-07-16 articles/formalism/article.tex
updates to introduction and revision to figures (removed NSIDE and corrected lmax numbers)

\author[]{Aditya Rotti and}
\author[]{Kevin Huffenberger}
\affiliation[]{Department of Physics, Florida State University, Keen Physics Building, Tallahassee, Florida 32306}
\emailAdd{adityarotti@gmail.com}
\emailAdd{khuffenberger@fsu.edu}

>>>>

\author[a,b]{Aditya Rotti}
\author[a]{and Kevin Huffenberger}
\affiliation[a]{Department of Physics, Florida State University, Keen Physics Building, Tallahassee, Florida 32306, USA}
\affiliation[b]{Jodrell Bank Center for Astrophysics, University of Manchester, Oxford Road, Manchester M13 9PL, UK}
\emailAdd{adityarotti@gmail.com}
\emailAdd{khuffenberger@fsu.edu}

### 2018-07-16 articles/formalism/sections/introduction.tex
updates to introduction and revision to figures (removed NSIDE and corrected lmax numbers)

\section{Introduction}
During recombination, the Cosmic Mircowave Background undergoes Thomson scattering that leaves it with $\sim 5$ percent linear polarization.  The polarization signal contains information about the plasma velocity and provides cosmological constraints independent from the signal in temperature \citep{1997NewA....2..323H}.  The standard analysis technique converts the Stokes parameters of polarization into scalar ($E$) and pseudo-scalar ($B$) modes, which are easier to compare to theory.  The pseudo-scalar $B$ mode is particularly important because it cannot be generated by primordial scalar pertubations.  Thus any $B$-modes that appear must be from:
% Need citations for all...
(1) primordial tensor perturbations (gravitational waves), which reveal the energy scale of inflation
\citep{1997PhRvD..56..596H,1997PhRvL..78.2054S};
(2) lensing of $E$-modes, which helps to reveal the distribution of matter in the universe;
(3) foregrounds (especially Galactic synchrotron and dust emission)
\citep{2016A&A...586A.133P};
(4) unknown systematic problems in the data;
(5) unknown, exotic phenemena like cosmic birefringence or primordial magnetic fields
\citep{1996ApJ...469....1K,1999PhRvL..83.1506L,2004ApJ...616....1C,2014MNRAS.438.2508P}.

The formalism for converting the Stokes parameters to scalar quantities is well established \citep{1997PhRvD..55.7368K,1997PhRvD..55.1830Z}.  The reason we take a detailed look at it here is that with renewed focus on foreground contamination to the $B$-mode signal, we wanted to gain an intuition for what $E/B$ mode patterns arise from physical polarized structures in the galaxy.  We also wanted to gain insight into the problem of masking and the non-locality of the $E/B$ fields compared to the Stokes parameters.

The spin-0 scalar $E$/$B$ modes relate to the spin-2 complex Stokes parameters via the spin-raising and -lowering operators ($\eth^2,\bar \eth^2$), which are derivatives evaluated locally.  However, in practice we compute $E$/$B$ modes to a specified band limit, and this makes them non-local functions of the polarization field.  In other words, the $E$/$B$ modes at a point can get contributions from all over the sky.

(stuff about foregrounds, masking, ambiguous modes with references)

Zaldarriaga explored the spatial real-space kernels in the flat-sky approximation \citep{Zaldarriaga2001a}.

This paper is organized in the following manner: In \sec{sec:pol-primer} we present a primer on the description of CMB polarization on the sphere. Here we begin with a heuristic argument that makes transparent the real space construction of E/B modes on the sphere. We discuss the standard harmonic space procedures for this operation. Finally we introduce the matrix-vector notation which yields a more concise description of the harmonic space procedures. In \sec{sec:qu2eb} and \sec{sec:eb2qu} we derive and discuss the real space operators that transform Q/U to E/B and vice versa. In \sec{sec:visualize_operator} we evaluate these real space operators and present visualizations of these functions. In \sec{sec:purify_stokes_qu} we derive a real space operator that decomposes the Stokes Q/U  parameters into components that correspond to E and B modes respectively and present its visualizations.  In \sec{sec:radial_locality} we study the locality of the real space operators and explore its band limit dependence. In \sec{sec:generalized_operators} we present a systematic way to generalize the real space operators by varying the non-locality parameter and discuss the connection to the spin raising and lower operators. We conclude with a summary and discuss the scope of this new method of analyzing CMB polarization in \sec{sec:discussion}.

>>>>

\section{Introduction}
During recombination, the Cosmic Mircowave Background undergoes Thomson scattering that leaves it with $\sim 5$ percent linear polarization.  The polarization signal contains information about the plasma velocity and provides cosmological constraints independent from the signal in temperature anisotropies \citep{1997NewA....2..323H}.  The standard analysis technique involves translating the Stokes parameters of polarization into scalar ($E$) and pseudo-scalar ($B$) modes, since the statistics of these coordinate independent scalar fields are predicted by theory.

The dominant contribution to the $E$-modes of polarization are sourced by primordial scalar perturbations, however these do not generate B-modes of polarization at first order. The $B$-modes of CMB polarization can be generated by various sources:
% Need citations for all...
(1) primordial tensor perturbations (gravitational waves)%, which reveal the energy scale of inflation
\citep{1997PhRvD..56..596H,1997PhRvL..78.2054S};
(2) weak gravitational lensing of $E$-modes resulting as a consequence of photon deflections induced by intervening large scale structures;
(3) foregrounds (especially Galactic synchrotron and dust emission)
\citep{2016A&A...586A.133P};
(4) unknown systematic problems in the data;
(5) unknown, exotic phenemena like cosmic birefringence or primordial magnetic fields
\citep{1996ApJ...469....1K,1999PhRvL..83.1506L,2004ApJ...616....1C,2014MNRAS.438.2508P}.
 The pseudo-scalar $B$-modes particularly on large angular scales ($\sim 2$ degrees) are expected to have a significant contribution from primordial tensor perturbations generated during inflation. Hence a measurement of the large angular scale B-modes will yield information about the statistical properties of tensor perturbations which will finally lead to important constraints on models of inflation \cite{?}. The measurement of $B$-modes on small angular scales ($\sim$ few arcminute) will yield information on the clustering of matter across cosmic ages \cite{?}.

The primary aim of the current CMB experiments is to make precise measurements of the CMB polarizations. While the E-modes of polarization have been measured reasonably well by a number of experiments \cite{? }, the ongoing and future CMB experiments aim to measure the B-modes of CMB polarization with unprecedented accuracy and angular resolution. The detectors are swiftly approaching the desired sensitivities enabling us to in principle measure tiny $B$-mode signals ($r\gtrsim 0.001$). However B-mode generated by galactic foreground are expected to be a few orders of magnitude higher than $B$-mode amplitude measurable by these detectors. Precise modelling and subtraction of this large foreground contribution poses a major challenge for a robust unravelling of the minuscule $B$-mode signal.

The formalism for converting the Stokes parameters to scalar quantities is well established \citep{1997PhRvD..55.7368K,1997PhRvD..55.1830Z}. The spin-0 scalar $E$/$B$ modes relate to the spin-2 complex Stokes parameters via the spin-raising and -lowering operators ($\eth^2,\bar \eth^2$), which are derivatives evaluated locally.  However, in practice the $E$/$B$ modes are computed to some specified band limit and this makes them non-local functions of the polarization field.  In other words, band limited $E$/$B$ modes evaluated at a point receive contributions from all over the sky. In this work we aim to gain real space insights into the non-locality of the $E/B$ fields compared to the Stokes parameters. With renewed focus on foreground contamination to the $B$-mode signal we aim to gain an intuition for $E/B$ mode patterns resulting from physical polarized structures in the galaxy. These real space insights may potentially yield new ideas for minimizing foreground contamination which may not be possible using conventional approaches.

Some of the ideas presented in this work bear resemblance with those in \citep{Zaldarriaga2001a}, however the approach adopted here is more rigorous and consequently there are updates to interpretation of the results and substantial differences in details.

This paper is organized in the following manner: In \sec{sec:pol-primer} we present a primer on the description of CMB polarization on the sphere. Here we begin with a heuristic argument that makes transparent the real space construction of E/B modes on the sphere. We discuss the standard harmonic space procedures for this operation. Finally we introduce the matrix-vector notation which yields a more concise description of the harmonic space procedures. In \sec{sec:qu2eb} and \sec{sec:eb2qu} we derive and discuss the real space operators that transform Q/U to E/B and vice versa. In \sec{sec:visualize_operator} we evaluate these real space operators and present visualizations of these functions. In \sec{sec:purify_stokes_qu} we derive a real space operator that decomposes the Stokes Q/U  parameters into components that correspond to E and B modes respectively and present its visualizations.  In \sec{sec:radial_locality} we study the locality of the real space operators and explore its band limit dependence. In \sec{sec:generalized_operators} we present a systematic way to generalize the real space operators by varying the non-locality parameter and discuss the connection to the spin raising and lower operators. In \sec{sec:pol_filaments} we discuss $E/B$ mode signatures of realistic filaments. We conclude with a summary and discuss the scope of this new method of analyzing CMB polarization in \sec{sec:discussion}.
